One-armed spirals in locally isothermal, radially structured self-gravitating discs

### 2014-11-17 paper/results_3d.tex
minor updates to tex

The 3D disc has radial size
$[r_\mathrm{min},r_\mathrm{max}]=[0.4,10]R_0$ and vertical extent  
$n_H=2$ scale-heights at $R=R_0$. The resolution is $N_r\times N_\theta\times
N_\phi=256\times32\times512$. This corresponds to about $4$ cells per
$H$ in radius. Because of the reduced resolution 
compared to 2D, we use a smooth perturbation by setting
$\delta = 10^{-3}$ and $M=1$ in Eq. \ref{randpert}. This corresponds
to a single $m=1$ disturbance in $R\in[R_1,R_2]$, which
eventually dominates in 2D simulations.

The 3D discs are initialised in approximate equilibrium only, so we
first evolve the disc without perturbations using  
$(\lmax,\mmax)=(32,0)$ up to $t=10P_0$, during which 
meridional velocities are damped out. We then restart the simulation
with the above perturbation and $(\lmax,\mmax)=(32,32)$, and damp
merdional velocities near the radial boundaries.

Fig. \ref{3d_ampmax} plots the amplitude of the $m=1$ spirals measured
in the ZEUS-MP and PLUTO runs.  
%why no high m
                                %modes as in 2D for iso disc?
                                %resolution? 
In the ZEUS-MP run we observe disturbances developed near
the inner boundary, which is probably responsible for the initial growth
for $t<70P_0$. This is likely a numerical artifact and effectively seeds
the ZEUS-UP simulation with a larger initial perturbation. Results
from ZEUS-MP are therefore off-set from PLUTO by $\sim50P_0$ However,
once the coherent $m=1$ spiral begins to grow ($t\gtrsim 100P_0$), 
we measure similar growth rates in both codes: 
\begin{align*}
  &\gamma \simeq 0.0073\Omega_k(R_0) \quad\quad \mathrm{PLUTO},\\
  &\gamma \simeq 0.0084\Omega_k(R_0) \quad\quad \varA{ZEUS-MP}.
\end{align*}
Both are smaller than the 2D simulations and due to the
low resolutions adopted and/or the small softening length employed in
2D.  Fig. \ref{3d_ampmax} also show results from strictly isothermal
simulations, which display no growth compared to cases with a 
temperature gradient. This confirms the temperature gradient effect is
the same in 3D.

@@

Visualisations of the 3D simluations are shown in   
Fig. \ref{3d_prelim} for the disc midplane and near the upper disc
boundary. The snapshots are chosen when the one-armeded spirals in the two codes
have reached similar amplitudes. The agreement between the codes,
as well as with the 2D simulations, is satisfactory. Away from the
disc midplane, a coherent spiral arm is not prominent, but the
disturbance is radially more global, extending up to $R\sim 4R_0$.   
%implications?

>>>>

The 3D disc has radial size
$[r_\mathrm{min},r_\mathrm{max}]=[0.4,10]R_0$ and vertical extent  
$n_H=2$ scale-heights at $R=R_0$. The resolution is $N_r\times N_\theta\times
N_\phi=256\times32\times512$, or about $4$ cells per
$H$. Because of the reduced resolution 
compared to 2D, we use a smooth perturbation by setting
$\delta = 10^{-3}$ and $M=1$ in Eq. \ref{randpert}. This corresponds
to a single $m=1$ disturbance in $R\in[R_1,R_2]$, which
eventually dominates in 2D simulations.

The 3D discs are initialised in approximate equilibrium only, so we
first evolve the disc without perturbations using  
$(\lmax,\mmax)=(32,0)$ up to $t=10P_0$, during which 
meridional velocities are damped out. We then restart the simulation
with the above perturbation and $(\lmax,\mmax)=(32,32)$, and damp
meridional velocities near the radial boundaries.

Fig. \ref{3d_ampmax} plots the evolution of the $m=1$ spiral amplitudes measured
in the ZEUS-MP and PLUTO runs. We also ran simulations
with a strictly isothermal equation of state ($q=0$), which display no
growth compared to that with a temperature gradient.  This confirms
the temperature gradient effect is the same in 3D. 
%why no high m
                                %modes as in 2D for iso disc?
                                %resolution? 
In the ZEUS-MP run we observe disturbances developed near
the inner boundary, which is probably responsible for the initial growth
for $t<70P_0$. This is likely a numerical artifact and effectively seeds
the ZEUS-UP simulation with a larger initial perturbation. Results
from ZEUS-MP are therefore off-set from PLUTO by $\sim50P_0$. However,
once the coherent $m=1$ spiral begins to grow ($t\gtrsim 100P_0$), 
we measure similar growth rates in both codes: 
\begin{align*}
  &\gamma \simeq 0.0073\Omega_k(R_0) \quad\quad \mathrm{PLUTO},\\
  &\gamma \simeq 0.0084\Omega_k(R_0) \quad\quad \varA{ZEUS-MP}.
\end{align*}
Both are smaller than the 2D simulations, because of to the
low resolutions adopted in 3D and/or the small softening length employed in
2D. 
%Fig. \ref{3d_ampmax} also show results from strictly isothermal
%simulations, which display no growth compared to cases with a 
%temperature gradient.

@@

Visualisations of the 3D simluations are shown in   
Fig. \ref{3d_prelim} for the disc midplane and near the upper disc
boundary. The snapshots are chosen when the one-armed spirals in the two codes
have reached comparable amplitudes. The agreement between the codes,
as well as with the 2D simulations, is satisfactory. The largest
spiral amplitude is found in $R\in[R_1,R_2]$ independent of
height. However, notice the spiral pattern extends into the
non-self-gravitating outer disc ($R>R_2$) at $z\sim 2H$, i.e. 
the disturbance becomes more global away from the midplane.

%Away from the
%disc midplane, a coherent spiral arm is not prominent, but the
%disturbance is radially more global, extending up to $R\sim 4R_0$.

%implications?
